Drought-sensitive trees survive on deep-water access in a tropical forest community

### 2020-08-30 manuscript.rmd
Equation alignment changed

---
title: "Drought-sensitive trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

\begin{equation}
\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
(\#eq:water-balance)
\end{equation}
where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step. $z$ represents the depths of 15 exponentially distributed soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).

@@

### Model evaluation
\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018). We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe ($\theta_{0-0.15 m}$). This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig. S\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 g cm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot, converted to VWC units assuming a bulk density of 0.8 g cm^-3^ ($\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019). Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data as that for the vertical probe. However, for VWC calibration we used data from only the vertical TDR probe ($\theta_{0-0.15 m}$) to capture the long-term dynamics and the plot-wide measurements to capture the dynamics by depth. Here, in general, our interest was to capture the soil moisture dynamics by depth, so we performed model calibration in a two step process: first, we prioritized model fit with $\theta_{0.1m}$. We chose ensembles that minimized root mean square error (RMSE) between $\theta_{0.1m}$ and W~liq,`r soil.depths[8]`m~ (~300 ensembles with standaridised RMSE (range 0-1) across ensembles < 0.2). Second, we calculated an objective function for each of the 5000 ensembles by equally weighing standardized RMSE between observations and simulations across all fluxes, and chose 100 ensembles, within the 300 that best-fit $\theta_{0-0.15 m}$, that had the lowest values of the objective function. Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function can be described as

\begin{equation}
Objective Function = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
(\#eq:obj-func)
\end{equation}
where $W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m. See Table \@ref(tab:param) for the range of best-fit values for different parameters. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

## Model for Effective Rooting Depths (ERD)
\
Water is a key limiting resource for tree growth. Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:

@@

\
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot and VPD (kPa; $\overline{GPP} = `r vpd.cf[1]` + `r vpd.cf[2]`VPD + `r vpd.cf[3]`VPD^2 + error = StdVPD$; 2012-2017; Fig. S\@ref(fig:gpp-vpd)) and used standardized $\overline{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs. \@ref(eq:M2), \@ref(eq:M3) \@ref(eq:M5) and \@ref(eq:M6). Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the impact of VPD may as well be a combined impact of temperature, humidity, and radiation.

\
To explore the effect of species seasonality in leaf cover on growth, we assumed of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless). For evergreen species we assumed $LeafCover$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LeafCover$ for the rest of the year as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (Fig. S\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardized K within species ($Std.K$; range 0-1) in Eqs. \@ref(eq:M1)- \@ref(eq:M6).

\
For the combination of each model \@ref(eq:M1)- \@ref(eq:M6), species and 100 hydrological realizations, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$. For each combination of model-species-hydrological realization we retained only one depth with the least AIC score (the best-fit), and within those depths retained the ones that had a positive corresponding Pearson's $\rho$ and *R*^2^ >= 0.5 as a valid set. For each model-species combination, we defined effective rooting depth (ERD) as the median of these subset of depths and calculated standard error.

@@

# Supplementary Information

@@

Among the `r length(sp.exp.param$sp)` species with observed data on ($K_{leaf}$ vs. $\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs. \@ref(eq:b-mod) and \@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~. We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj. *R*^2^ = `r disc.r2`; *P* `r disc.p`, *n* = `r disc.n`). We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj. *R*^2^ = `r lma.r2`; *P* `r lma.p`, *n* = `r lma.n`). We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.

@@

```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for ", length(sp.exp.param$sp)," species (Eq. \\@ref(eq:vc-exp)) vs. those that were predicted from models based on species 'soft-traits', $WSG$ and $LMA$. (Eqs. \\@ref(eq:a-mod) & \\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *P*-values are given in inset.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
```

>>>>

---
title: "Drought-sensitive trees survive on deep-water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

@@

\begin{equation}
\begin{split}
\Delta W_{can} + \Delta W_{sfc} + \sum_{z=1}^{15} \Delta W_{liq,z} + \Delta W_{a} = \left(q_{rain} - E_{v} - E_{g} - q_{over} - q_{h2osfc} - q_{drai}\right)\Delta t
\end{split}
(\#eq:water-balance)
\end{equation}

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step. $z$ represents the depths of 15 exponentially distributed soil layers: $\simeq$ 0.01, 0.03, 0.06, 0.12, 0.21, 0.37, 0.62, 1, 1.7, 2.9, 4.7, 7.8, 13, 21 and 35 (m).

@@

### Model evaluation
\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017; see [SI](#SIData)), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) soil volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018). We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical time domain reflectometry (TDR) probe ($\theta_{0-0.15 m}$). This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig. S\@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 g cm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). We also leveraged existing opportunistic dry season gravimetric measurements by depth for the entire 50-ha plot, converted to VWC units assuming a bulk density of 0.8 g cm^-3^ ($\theta_{0.15 m}$, $\theta_{0.4m}$ and $\theta_{1 m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019). Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1 m and we applied the same calibration equation to these data as that for the vertical probe. However, for VWC calibration we used data from only the vertical TDR probe ($\theta_{0-0.15 m}$) to capture the long-term dynamics and the plot-wide measurements to capture the dynamics by depth. Here, in general, our interest was to capture the soil moisture dynamics by depth, so we performed model calibration in a two step process: first, we prioritized model fit with $\theta_{0.1m}$. We chose ensembles that minimized root mean square error (RMSE) between $\theta_{0.1m}$ and W~liq,`r soil.depths[8]`m~ (~300 ensembles with standaridised RMSE (range 0-1) across ensembles < 0.2). Second, we calculated an objective function for each of the 5000 ensembles by equally weighing standardized RMSE between observations and simulations across all fluxes, and chose 100 ensembles, within the 300 that best-fit $\theta_{0-0.15 m}$, that had the lowest values of the objective function. Let $f(obs,~sim)$ be RMSE between observations and simulations standardized across ensembles (range 0-1), based on monthly sums for $ET$ and $q_{run}$, respectively, and daily averages for $\theta$s, then the objective function ($\overline{O}$) can be described as

\begin{equation}
\overline{O} = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\overline{z}}}) + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1 m}, W_{liq,~`r soil.depths[8]`m})\right)
(\#eq:obj-func)
\end{equation}

where $W_{liq,~\overline{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m. See Table \@ref(tab:param) for the range of best-fit values for different parameters. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

## Model for Effective Rooting Depths (ERD)
\
Water is a key limiting resource for tree growth. Vapor pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_{leaf}$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)-\@ref(eq:M6)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied or additive effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardized to range from 0-1; see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows:

@@

\
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between mean daily values by month of gross primary productivity (gC m^2^ day^-1^) we collected from the eddy flux tower located 75 m from the 50-ha plot and VPD (kPa; $\overline{GPP} = `r vpd.cf[1]` + `r vpd.cf[2]`VPD + `r vpd.cf[3]`VPD^2 + error = StdVPD$; 2012-2017; Fig. S\@ref(fig:gpp-vpd)) and used standardized $\overline{GPP}$ to represent the impact of VPD on growth (hereafter, $StdVPD$; range 0-1) in Eqs. \@ref(eq:M2), \@ref(eq:M3), \@ref(eq:M5) and \@ref(eq:M6). Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at our study site, the impact of VPD may as well be a combined impact of temperature, humidity, and radiation.

\
To explore the effect of species seasonality in leaf cover on growth, we assumed of species-specific seasonal curves for Leaf Cover Fraction ($LeafCover$, range 0-1, unitless). For evergreen species we assumed $LeafCover$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LeafCover$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LeafCover$ for the rest of the year as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LeafCover$ = 1 beginning DOY = 300 and obtained a $LeafCover$ by DOY curve (Fig. S\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used standardized K within species ($Std.K$; range 0-1) in Eqs. \@ref(eq:M1)-\@ref(eq:M6).

\
For the combination of each model \@ref(eq:M1)-\@ref(eq:M6), species and 100 hydrological realizations, we obtained AIC, goodness-of-fit (*R*^2^) and Pearson's $\rho$ for each soil depth $z$. For each combination of model-species-hydrological realization we retained only one depth with the least AIC score (the best-fit), and within those depths retained the ones that had a positive corresponding Pearson's $\rho$ and *R*^2^ >= 0.5 as a valid set. For each model-species combination, we defined effective rooting depth (ERD) as the median of these subset of depths and calculated standard error.

@@

\newpage

# Supplementary material {-}

@@

Among the `r length(sp.exp.param$sp)` species with observed data on ($K_{leaf}$ vs. $\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs. \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~. We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj. *R*^2^ = `r disc.r2`; *P* `r disc.p`, *n* = `r disc.n`). We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj. *R*^2^ = `r lma.r2`; *P* `r lma.p`, *n* = `r lma.n`). We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.

@@

```{r ab-fit, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', dpi = 96, out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-ab-fit}Parameters $A$ (panel A) & $B$ (panel B) of leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for ", length(sp.exp.param$sp)," species (Eq. \\@ref(eq:vc-exp)) vs. those that were predicted from models based on species 'soft-traits', $WSG$ and $LMA$. (Eqs. \\@ref(eq:a-mod)-\\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *P*-values are given in inset.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/A_B_data_vs_model.jpeg"))
```
